Starcode: sequence clustering based on all-pairs search

### 2015-01-20 doc/document.tex
Manuscript updated.

\subsection{Clustering}
The default clustering algorithm of starcode is designed to correct
sequencing error. This method uses message passing \citep{mackay}
to identfy and count ``canonical'' sequences (also referred to as
centroids in the clustering terminology). Each sequence transfers its
read count to its closest $\tau$-match provided the latter has at least 5
times more counts. If the condition is not met, the transfer does
not take place. If the sequence has several equally close
$\tau$-matches, the counts are split equally among them. The
process is repeated recursively, starting from sequences with
lowest read count. The sequences with a positive read count at
the end of the process are considered canonical. Clusters consist
of all the sequences transferring their read counts to the same
canonical sequence (sequence transferring their read counts to
different canonicals are discarded). Note that the radius of the
clusters can be higher than the maximum distance used for matching.

Since no sequencing technology has an error rate higher than 20\%,
it is expected that sequences appearing from sequencing errors will
always have 5 times or lower read count than the canonical sequence.
Otherwise, sequences are more likely unrelated, or both are derived
from the same canonical sequence.

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. File formats compatible
with starcode are raw sequence, raw sequence with count, FASTA or
FASTQ (in which case starcode ignores the quality). Starcode either
returns detailed information of the clustering results, i.e. canonical
sequences, cluster sizes and the complete list of their constituent
sequences. Alternatively, only the canonical sequences are
printed, which is useful to filter out redundant
sequences from input files. By default, clustering is performed under
the assumption that divergence occurs from experimental errors
(sequencing errors, PCR mutations \textit{etc}.) and a more general
algorithm is also available for other clustering problems (an example
of which is given in section \ref{section:motifs}).

We show the basic performance and scalability of starcode on a
dataset of pseudorandom sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. In each
test, only one parameter is modified while the others are kept constant.

Figure \ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic
scale the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (the
naive implementation of all-pairs search). Note that the sequences of
this dataset have no match, see section \ref{section:benchmark} for
an evaluation of the performance on more realistic datasets.
Figure \ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing number
of threads in Figure \ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is due to hardware limitations, i.e. there is insufficient
memory bandwidth to satisfy the increased demand of parallel memory accesses.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets (Figure \ref{fig:benchmark}a). We also observed that
slidesort found 5-10\% less 3-matches than starcode on all the
datasets (Figure \ref{fig:benchmark}b). We were surprised by 
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons are feasible
(more information is available on the starcode repository the, see
http://github.com/gui11aume/starcode/misc).

>>>>

\subsection{Clustering}
The default clustering algorithm of starcode is designed to correct
sequencing error. This method uses message passing \citep{mackay}
to identify and count ``canonical'' sequences (also referred to as
centroids in the clustering terminology). By default, each sequence transfers its
read count to its closest $\tau$-match provided the latter has at least 5
times more counts. If the condition is not met, the transfer does
not take place. If the sequence has several equally close
$\tau$-matches, the counts are split equally among them. The
process is repeated recursively, starting from sequences with
lowest read count. The sequences with a positive read count at
the end of the process are considered canonical. Clusters consist
of all the sequences transferring their read counts to the same
canonical sequence (sequence transferring their read counts to
different canonicals are discarded). Note that the radius of the
clusters can be higher than the maximum distance used for matching.

Since no sequencing technology has an error rate higher than 20\%,
it is expected that sequences appearing from sequencing errors will
always have 5 times or lower read count than the canonical sequence.
Otherwise, sequences are more likely unrelated, or both are derived
from the same canonical sequence. This behavior can be modified with
the command-line option \emph{cluster-ratio} to allow for a more
flexible or more strict clustering, e.g. to cluster unique input
sequences together, \emph{cluster-ratio} must be set to 1.

@@

\subsection{Presentation and basic performance}
\label{section:performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. File formats compatible
with starcode are raw sequence, raw sequence with count, FASTA or
FASTQ (in which case starcode ignores the quality). Starcode either
returns detailed information of the clustering results, i.e. canonical
sequences, cluster sizes and the complete list of their constituent
sequences. Alternatively, only the canonical sequences are
printed, which is useful to filter out redundant
sequences from input files. By default, clustering is performed under
the assumption that divergence occurs from experimental errors
(sequencing errors, PCR mutations \textit{etc}.) and a more general
algorithm is also available for other clustering problems (an example
of which is given in section \ref{section:motifs}).

We show the basic performance and scalability of starcode on a
dataset of pseudorandom sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. In each
test, only one parameter is modified while the others are kept
constant. Since the clustering step does not require additional memory
allocation and is significantly faster than all-pairs search, the
performance results presented in Sections \ref{section:performance} and
\ref{section:benchmark} apply for both message-passing and spheres
clustering algorithms.

Figure \ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic
scale the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (the
naive implementation of all-pairs search). Note that the sequences of
this dataset have no match, see section \ref{section:benchmark} for
an evaluation of the performance on more realistic datasets.
Figure \ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing number
of threads in Figure \ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources; the first is 
that the input reading and clustering steps are brief but not parallel,
the second is due to hardware limitations, i.e. there is insufficient
memory bandwidth to satisfy the increased demand of parallel memory accesses.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets (Figure \ref{fig:benchmark}a). We also observed that
slidesort found 5-10\% less 3-matches than starcode on all the
datasets (Figure \ref{fig:benchmark}b). We were surprised by 
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons are feasible
(more information is available on the starcode repository, see
http://github.com/gui11aume/starcode/tree/master/misc).
